---
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    code_folding: hide
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: spacelab
  pdf_document:
      toc: true
  knitrBootstrap::bootstrap_document:
    theme.chooser: TRUE
    highlight.chooser: TRUE
---


Basic `r software` results exploration
================================

Comparison: `r project`

# Introduction

This report is meant to help explore `r software` `r if(software == 'edgeR') citep(bib[[c('edgeR1', 'edgeR2', 'edgeR6')]]) else if (software == 'DESeq2') citep(bib[['DESeq2']]) else citep(bib[['other']])` results and was generated using the `regionReport` `r citep(bib[['regionReport']])` package. While the report is rich, it is meant to just start the exploration of the results and exemplify some of the code used to do so. If you need a more in-depth analysis for your specific data set you might want to use the `customCode` argument. This report is based on the vignette of the `DESeq2` `r citep(bib[['DESeq2']])` package which you can find [here](http://www.bioconductor.org/packages/DESeq2).

## Code setup

This section contains the code for setting up the rest of the report.

```{r docSetup, bootstrap.show.code = FALSE, dev = device, bootstrap.show.message=FALSE}
## knitrBoostrap and device chunk options
library('knitr')
opts_chunk$set(bootstrap.show.code = FALSE, dev = device, crop = NULL)
if(!outputIsHTML) opts_chunk$set(bootstrap.show.code = FALSE, dev = device, echo = FALSE)
```

```{r setup, bootstrap.show.message=FALSE, message=FALSE}
#### Libraries needed

## Bioconductor
library('DESeq2')
if(isEdgeR) library('edgeR')

## CRAN
library('ggplot2')
if(!is.null(theme)) theme_set(theme)
library('knitr')
if(is.null(colors)) {
    library('RColorBrewer')
}
library('pheatmap')
library('EnhancedVolcano')
library('DT')
library('sessioninfo')

#### Code setup

## For ggplot
res.df <- as.data.frame(res)

## Sort results by adjusted p-values
ord <- order(res.df$padj, decreasing = FALSE)
res.df <- res.df[ord, ]
features <- rownames(res.df)
res.df <- cbind(data.frame(Feature = features), res.df)
rownames(res.df) <- NULL
```

# PCA

The below plots show the first two principal components that explain the variability in the data using the regularized log (`rlog`) and variance stabilizing transformed (`vsd`) count data respectively. If you are unfamiliar with principal component analysis, you might want to check the [Wikipedia entry](https://en.wikipedia.org/wiki/Principal_component_analysis) or this [interactive explanation](http://setosa.io/ev/principal-component-analysis/). 

```{r 'PCA_rlog'}
## Transform count data
rld <- tryCatch(rlog(dds), error = function(e) { rlog(dds, fitType = 'mean') })

## Get percent of variance explained
data_pca_rld <- plotPCA(rld, intgroup=intgroup, returnData=TRUE) 
percentVar_rld <- round(100 * attr(data_pca_rld, "percentVar"))

ggplot(data_pca_rld, aes(PC1, PC2, color=condition, shape=col.labels)) +
    scale_shape_manual(values=1:length(col.labels)) +
    geom_point(size=3) +
    xlab(paste0("PC1: ",percentVar_rld[1],"% variance")) +
    ylab(paste0("PC2: ",percentVar_rld[2],"% variance")) +
    coord_fixed() + 
    theme(legend.box = "horizontal") +
    ggtitle("PC1 vs PC2: 500 genes", subtitle = "regularized log transformed counts")
```

In this case (`rlog` count data), the first and second principal component explain `r percentVar_rld[1]` and `r percentVar_rld[2]` percent of the variance respectively. 

```{r 'PCA_vsd'}
## Transform count data
# vsd <- tryCatch(vst(dds, blind=FALSE))
try.vst <- try(
  vsd <- vst(dds, blind=FALSE)
)
if (class(try.vst) == "try-error") {
  vsd <- varianceStabilizingTransformation(dds, blind=FALSE)
}
## Get percent of variance explained
data_pca_vsd <- plotPCA(vsd, intgroup=intgroup, returnData=TRUE) 
percentVar_vsd <- round(100 * attr(data_pca_vsd, "percentVar"))

ggplot(data_pca_vsd, aes(PC1, PC2, color=condition, shape=col.labels)) +
    scale_shape_manual(values=1:length(col.labels)) +
    geom_point(size=3) +
    xlab(paste0("PC1: ",percentVar_vsd[1],"% variance")) +
    ylab(paste0("PC2: ",percentVar_vsd[2],"% variance")) +
    coord_fixed() + 
    theme(legend.box = "horizontal") +
    ggtitle("PC1 vs PC2: 500 genes", subtitle = "variance stabilizing transformed counts")
```

In this case (`vsd` count data), the first and second principal component explain `r percentVar_vsd[1]` and `r percentVar_vsd[2]` percent of the variance respectively.

# Sample-to-sample distances

```{r 'sampleDist'}
## Obtain the sample euclidean distances
sampleDists <- dist(t(assay(rld)))
sampleDistMatrix <- as.matrix(sampleDists)
## Add names based on intgroup
rownames(sampleDistMatrix) <- apply(as.data.frame(colData(rld)[, intgroup]), 1,
    paste, collapse = ' : ')
colnames(sampleDistMatrix) <- NULL

## Define colors to use for the heatmap if none were supplied
if(is.null(colors)) {
    colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
}

## Make the heatmap
pheatmap(sampleDistMatrix, clustering_distance_rows = sampleDists,
    clustering_distance_cols = sampleDists, color = colors)
```

This plot shows how samples are clustered based on their euclidean distance using the regularized log transformed count data. This figure gives an overview of how the samples are hierarchically clustered. It is a complementary figure to the PCA plot.

# MA plots

This section contains three MA plots (see [Wikipedia](https://en.wikipedia.org/wiki/MA_plot)) that compare the mean of the normalized counts against the log fold change. They show one point per feature. The points are shown in red if the feature has an adjusted p-value less than `alpha`, that is, the statistically significant features are shown in red.

```{r 'MAplotalpha'}
## MA plot with alpha used in DESeq2::results()
plotMA(res, alpha = metadata(res)$alpha, main = paste('MA plot with alpha =',
    metadata(res)$alpha))
```

This first plot shows uses `alpha` = `r metadata(res)$alpha`, which is the `alpha` value used to determine which resulting features were significant when running the function `DESeq2::results()`.


```{r 'MAplotalphaHalf'}
## MA plot with alpha = 1/2 of the alpha used in DESeq2::results()
plotMA(res, alpha = metadata(res)$alpha / 2,
    main = paste('MA plot with alpha =', metadata(res)$alpha / 2))
```

This second MA plot uses `alpha` = `r metadata(res)$alpha / 2` and can be used agains the first MA plot to identify which features have adjusted p-values between `r metadata(res)$alpha / 2` and `r metadata(res)$alpha`.

```{r 'MAplotalpha-nBest'}
## MA plot with alpha corresponding to the one that gives the nBest features
nBest.actual <- min(nBest, nrow(head(res.df, n = nBest)))
nBest.alpha <- head(res.df, n = nBest)$padj[nBest.actual]
plotMA(res, alpha = nBest.alpha * 1.00000000000001,
    main = paste('MA plot for top', nBest.actual, 'features'))
```

The third and final MA plot uses an alpha such that the top `r nBest.actual` features are shown in the plot. These are the features that whose details are included in the _top features_ interactive table.

# P-values distribution


```{r pvalueHistogram}
## P-value histogram plot
ggplot(res.df[!is.na(res.df$pvalue), ], aes(x = pvalue)) +
    geom_histogram(alpha=.5, position='identity', bins = 50) +
    labs(title='Histogram of unadjusted p-values') +
    xlab('Unadjusted p-values') +
    xlim(c(0, 1.0005))
```

This plot shows a histogram of the unadjusted p-values. It might be skewed right or left, or flat as shown in the [Wikipedia examples](https://en.wikipedia.org/wiki/Histogram#Examples). The shape depends on the percent of features that are differentially expressed. For further information on how to interpret a histogram of p-values check [David Robinson's post on this topic](http://varianceexplained.org/statistics/interpreting-pvalue-histogram/).

```{r pvalueSumm}
## P-value distribution summary
summary(res.df$pvalue)
```

This is the numerical summary of the distribution of the p-values.


```{r pvalueTable, results = 'asis'}
## Split features by different p-value cutoffs
pval_table <- lapply(c(1e-04, 0.001, 0.01, 0.025, 0.05, 0.1, 0.2, 0.3, 0.4, 0.5,
    0.6, 0.7, 0.8, 0.9, 1), function(x) {
    data.frame('Cut' = x, 'Count' = sum(res.df$pvalue <= x, na.rm = TRUE))
})
pval_table <- do.call(rbind, pval_table)
if(outputIsHTML) {
    kable(pval_table, format = 'markdown', align = c('c', 'c'))
} else {
    kable(pval_table)
}
```

This table shows the number of features with p-values less or equal than some commonly used cutoff values. 


# Adjusted p-values distribution

```{r padjHistogram}
## Adjusted p-values histogram plot
ggplot(res.df[!is.na(res.df$padj), ], aes(x = padj)) +
    geom_histogram(alpha=.5, position='identity', bins = 50) +
    labs(title=paste('Histogram of', elementMetadata(res)$description[grep('adjusted', elementMetadata(res)$description)])) +
    xlab('Adjusted p-values') +
    xlim(c(0, 1.0005))
```

This plot shows a histogram of the `r elementMetadata(res)$description[grep('adjusted', elementMetadata(res)$description)]`. It might be skewed right or left, or flat as shown in the [Wikipedia examples](https://en.wikipedia.org/wiki/Histogram#Examples). 

```{r padjSumm}
## Adjusted p-values distribution summary
summary(res.df$padj)
```

This is the numerical summary of the distribution of the `r elementMetadata(res)$description[grep('adjusted', elementMetadata(res)$description)]`.


```{r padjTable, results = 'asis'}
## Split features by different adjusted p-value cutoffs
padj_table <- lapply(c(1e-04, 0.001, 0.01, 0.025, 0.05, 0.1, 0.2, 0.3, 0.4, 0.5,
    0.6, 0.7, 0.8, 0.9, 1), function(x) {
    data.frame('Cut' = x, 'Count' = sum(res.df$padj <= x, na.rm = TRUE))
})
padj_table <- do.call(rbind, padj_table)
if(outputIsHTML) {
    kable(padj_table, format = 'markdown', align = c('c', 'c'))
} else {
    kable(padj_table)
}
```

This table shows the number of features with `r elementMetadata(res)$description[grep('adjusted', elementMetadata(res)$description)]` less or equal than some commonly used cutoff values. 



# Top features

This `r ifelse(outputIsHTML, 'interactive', '')` table shows the top `r nBest.actual` features ordered by their `r elementMetadata(res)$description[grep('adjusted', elementMetadata(res)$description)]`. `r ifelse(outputIsHTML, 'Use the search function to find your feature of interest or sort by one of the columns.', 'Since the report is in PDF format, only the top 20 features are shown.')`

```{r 'topFeatures', results = 'asis'}
## Add search url if appropriate
if(!is.null(searchURL) & outputIsHTML) {
    res.df$Feature <- paste0('<a href="', searchURL, res.df$Feature, '">',
        res.df$Feature, '</a>')
}
for(i in which(colnames(res.df) %in% c('pvalue', 'padj'))) res.df[, i] <- format(res.df[, i], scientific = TRUE)
if(outputIsHTML) {
    datatable(head(res.df, n = nBest), options = list(pagingType='full_numbers', pageLength=10, scrollX='100%'), escape = FALSE, rownames = FALSE) %>% formatRound(which(!colnames(res.df) %in% c('pvalue', 'padj', 'Feature')), digits)
} else {
    res.df_top <- head(res.df, n = 20)
    for(i in which(!colnames(res.df) %in% c('pvalue', 'padj', 'Feature'))) res.df_top[, i] <- round(res.df_top[, i], digits)
    kable(res.df_top)
}
```




# Count plots top features


This section contains plots showing the normalized counts per sample for each group of interest. Only the best `r nBestFeatures` features are shown, ranked by their `r elementMetadata(res)$description[grep('adjusted', elementMetadata(res)$description)]`. The Y axis is on the log10 scale and the feature name is shown in the title of each plot.


```{r 'plotCounts'}
plotCounts_gg <- function(i, dds, intgroup) {
    group <- if (length(intgroup) == 1) {
        colData(dds)[[intgroup]]
    } else if (length(intgroup) == 2) {
        lvls <- as.vector(t(outer(levels(colData(dds)[[intgroup[1]]]), 
            levels(colData(dds)[[intgroup[2]]]), function(x, 
                y) paste(x, y, sep = " : "))))
        droplevels(factor(apply(as.data.frame(colData(dds)[, 
            intgroup, drop = FALSE]), 1, paste, collapse = " : "), 
            levels = lvls))
    } else {
        factor(apply(as.data.frame(colData(dds)[, intgroup, drop = FALSE]), 
            1, paste, collapse = " : "))
    }
    data <- plotCounts(dds, gene=i, intgroup=intgroup, returnData = TRUE)
    ## Change in version 1.15.3
    ## It might not be necessary to have any of this if else, but I'm not
    ## sure that plotCounts(returnData) will always return the 'group' variable.
    if('group' %in% colnames(data)) {
        data$group <- group
    } else {
        data <- cbind(data, data.frame('group' = group))
    }
    ggplot(data, aes(x = group, y = count)) + geom_point() + ylab('Normalized count') + ggtitle(i) + coord_trans(y = "log10") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
}
for(i in head(features, nBestFeatures)) {
    print(plotCounts_gg(i, dds = dds, intgroup = intgroup))
}
```


`r ifelse(isEdgeR, '# edgeR specific plots', '')`

`r ifelse(isEdgeR, '## Biological coefficient of variation', '')`

```{r 'edgeR-BCV', eval = isEdgeR, echo = isEdgeR & outputIsHTML}
## Make BCV plot for edgeR results
plotBCV(dge)
```

`r ifelse(isEdgeR, 'This plot shows the feature-wise biological coefficient of variation (BCV) against the feature abundances in log2 counts per million. The plot shows the common, trended and feature-wise BCV estimates. If using _edgeR-robust_ only the trend and tagwise are shown. Check the [edgeR vignette](http://www.bioconductor.org/packages/edgeR) for further details regarding this plot.', '')`


`r ifelse(isEdgeR, '## MDS plot of distances', '')`

```{r 'edgeR-MDS', eval = isEdgeR, echo = isEdgeR & outputIsHTML}
## Make MDS plot for edgeR results
plotMDS(dge)
```

`r ifelse(isEdgeR, 'This plot is a multidimensional scaling plot of distances between feature expression profiles. It shows the names of the samples in a two-dimensional scatterplot such that the distances are approximately the log2 fold changes between samples. Check the [edgeR vignette](http://www.bioconductor.org/packages/edgeR) for further details regarding this plot.', '')`

```{r child = customCode, eval = hasCustomCode}
```

# Volcano plot

```{r volcanoPlot}
EnhancedVolcano(res, lab = rownames(res), x = 'log2FoldChange', y = 'padj', legendLabels = c('NS', expression(Log[2]~FC), "adj. p-value", expression(adj.~p-value~and~log[2]~FC)))
```

The above plot shows a volcano plot (see [Wikipedia](https://en.wikipedia.org/wiki/Volcano_plot_(statistics))) that compares the log2 fold change against the adjusted P value. The plot was created with [EnhancedVolcano](https://github.com/kevinblighe/EnhancedVolcano).


# Reproducibility

The input for this report was generated with `r software` `r if(software == 'edgeR') citep(bib[[c('edgeR1', 'edgeR2', 'edgeR6')]]) else if (software == 'DESeq2') citep(bib[['DESeq2']]) else citep(bib[['other']])` `r if(software == 'DESeq2') paste('using version', metadata(dds)$version)` and the resulting features were called significantly differentially expressed if their `r elementMetadata(res)$description[grep('adjusted', elementMetadata(res)$description)]` were less than `alpha` = `r metadata(res)$alpha`. This report was generated in path `r tmpdir` using the following call to `r ifelse(software != 'edgeR', 'DESeq2Report()', 'edgeReport()')`:

```{r thecall, echo=FALSE}
theCall
```

Date the report was generated:

```{r reproducibility1, echo=FALSE}
## Date the report was generated
Sys.time()
```

Wallclock time spent generating the report:

```{r reproducibility2, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits=3)
```

`R` session information.

```{r reproducibility3, echo=FALSE}
## Session info
options(width = 120)
session_info()
```

Pandoc version used: `r rmarkdown::pandoc_version()`.

# Bibliography

This report was created with `regionReport` `r citep(bib[['regionReport']])` using `rmarkdown` `r citep(bib[['rmarkdown']])` while `knitr` `r citep(bib[['knitr']])` and  `DT` `r citep(bib[['DT']])` were running behind the scenes. `pheatmap` `r citep(bib[['pheatmap']])` was used to create the sample distances heatmap. Several plots were made with `ggplot2` `r citep(bib[['ggplot2']])`. The volcano plot was create with `EnhancedVolcano` ([Blighe, Rana, and Lewis, 2018](https://github.com/kevinblighe/EnhancedVolcano)).

Citations made with `knitcitations` `r citep(bib[['knitcitations']])`. The [BibTeX](http://www.bibtex.org/) file can be found [here](`r paste0(output, '.bib')`).
```{r bibliography, results='asis', echo=FALSE, warning = FALSE}
## Print bibliography
bibliography()
```

